---
title: "RQ2: Behavioral Factors Associated with Abnormal Cholesterol"
author: "Zhan Zhao"
date: "2026-04-28"
output:
  word_document: default
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    code_folding: hide
---

# Research Question 
Behavioral Factors Associated with Abnormal Cholesterol, Among U.S. young adults aged 18–26 years 

# ::: {.hypothesis #label name="hypothesis name"}
 
**Null Hypothesis:** Smoking status, alcohol use, physical activity, and sedentary behavior  
 are not associated with abnormal cholesterol after adjustment for demographic and body weight characteristics.  

**Alternative Hypothesis:** One or more behavior factors are associatedwith abnormal cholesterol after adjustment  
 for demographic and body weight characteristics.
:::

# Initialize

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                        warning = TRUE,
                        message = TRUE,
                        fig.width = 8,
                        fig.height = 6
                        )
library(tidyr)
library(survey)
library(ggplot2)
library(srvyr)
library(dplyr)
library(haven)
library(janitor)
```



# Define Data Directories
```{r define-data-directories}

project.dir <- "/Users/zhaozhan/nhanes-abnormal-cholesterol-young-adults"
data.dir <- file.path(project.dir, "01 Data")
output.dir <- file.path(project.dir, "02 Output")
processed.dir <- file.path(data.dir, "processed")


nhanes.young <- readRDS(
    file.path(processed.dir, "nhanes_young_cholesterol.rds")
)

# Questionaire files for RQ2 behavioral factors
alq.file <- file.path(data.dir, "ALQ_L.xpt")
paq.file <- file.path(data.dir, "PAQ_L.xpt")
slq.file <- file.path(data.dir, "SLQ_L.xpt")
smq.file <- file.path(data.dir, "SMQ_L.xpt")

alq <- read_xpt(alq.file) %>% clean_names()
paq <- read_xpt(paq.file) %>% clean_names()
slq <- read_xpt(slq.file) %>% clean_names()
smq <- read_xpt(smq.file) %>% clean_names()

# Check variable names
names(alq)
names(paq)
names(slq)
names(smq)

```



# Load and Clean Data 
```{r merge-data}
# Merge the behavioral data with the main dataset 
nhanes.q2 <- nhanes.young |>
    left_join(
        alq |> select(seqn, alq111, alq121, alq170), 
        by = "seqn") |>
    left_join(
        paq |> select(seqn, pad790q, pad790u, pad800, pad810q, pad810u, pad820, pad680), 
        by = "seqn") |>
    left_join(
        slq |> select(seqn, sld012, sld013),
        by = "seqn") |>
    left_join(
        smq |> select(seqn, smq020, smq040 ),
         by = "seqn")

```

# Create Analytic Variables and Survey Design

```{r create-analytic-variables}
nhanes.q2 <- nhanes.q2 |>
    mutate(
        # smoking status
        smoking_status = case_when(
            smq020 == 2 ~ "Never smoker",
            smq020 == 1 & smq040 == 3 ~ "Former smoker",
            smq020 == 1 & smq040 %in% c(1,2) ~ "Current smoker",
            TRUE ~ NA_character_
        ),
        # alcohol use
        alcohol_12mo = case_when(
            alq111 == 2 ~ "Never drinker",
            alq121 == 0 ~ "No alcohol in past 12 months",
            alq121 %in% 1:10 ~ "Alcohol use in past 12 months",
            TRUE ~ NA_character_
        ),
        

    binge_30d = case_when(

      alq170 == 0 ~ "No binge drinking in past 30 days",

      alq170 >= 1 & alq170 <= 30 ~ "Any binge drinking in past 30 days",

      TRUE ~ NA_character_

    ),

    moderate_freq_week = case_when(

      pad790u == "D" ~ pad790q * 7,

      pad790u == "W" ~ pad790q,

      pad790u == "M" ~ pad790q / 4.345,

      pad790u == "Y" ~ pad790q / 52,

      TRUE ~ NA_real_

    ),

    vigorous_freq_week = case_when(

      pad810u == "D" ~ pad810q * 7,

      pad810u == "W" ~ pad810q,

      pad810u == "M" ~ pad810q / 4.345,

      pad810u == "Y" ~ pad810q / 52,

      TRUE ~ NA_real_

    ),

    moderate_min_week = moderate_freq_week * pad800,

    vigorous_min_week = vigorous_freq_week * pad820,

    total_pa_min_week = rowSums(

      cbind(moderate_min_week, vigorous_min_week),

      na.rm = TRUE

    ),

    pa_category = case_when(

      total_pa_min_week >= 150 ~ "Meets 150 min/week",

      total_pa_min_week < 150 ~ "Below 150 min/week",

      TRUE ~ NA_character_

    ),

    sedentary_cat = case_when(

      pad680 >= 480 & pad680 < 1440 ~ "High sedentary time",

      pad680 < 480 ~ "Lower sedentary time",

      TRUE ~ NA_character_

    ),

    sleep_cat = case_when(

      sld012 < 7 ~ "Short sleep",

      sld012 >= 7 & sld012 <= 9 ~ "Recommended sleep",

      sld012 > 9 ~ "Long sleep",

      TRUE ~ NA_character_
    )

 )

nhanes.q2.design <- nhanes.q2 |>
  as_survey_design(
    ids = sdmvpsu,
    strata = sdmvstra,
    weights = wtmec2yr,
    nest = TRUE
  )

```

```{r descriptive-table}

# smking status
q2.desc <- nhanes.q2.design |>
  group_by(smoking_status) |>
  summarise(
    prevalence = survey_mean(flag.any, na.rm = TRUE)
  )

print(q2.desc)
q2.alcohol <- nhanes.q2.design |>
  group_by(binge_30d) |>
  summarise(
    prevalence = survey_mean(flag.any, na.rm = TRUE)
  )

print(q2.alcohol)


q2.pa <- nhanes.q2.design |>
  group_by(pa_category) |>
  summarise(
    prevalence = survey_mean(flag.any, na.rm = TRUE)
  )

print(q2.pa)

q2.sedentary <- nhanes.q2.design |>
  group_by(sedentary_cat) |>
  summarise(
    prevalence = survey_mean(flag.any, na.rm = TRUE)
  )

print(q2.sedentary)
```
