---
title: "RQ2: Behavioral Factors Associated with Abnormal Cholesterol"
author: "Zhan Zhao"
date: "2026-04-28"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    code_folding: hide
  word_document: default
---

# Research Question 
Behavioral Factors Associated with Abnormal Cholesterol, Among U.S. young adults aged 18–26 years 

::: {.hypothesis #label name="hypothesis name"}
 
**Null Hypothesis:** Smoking status, alcohol use, physical activity, and sedentary behavior  
 are not associated with abnormal cholesterol after adjustment for demographic and body weight characteristics.  

**Alternative Hypothesis:** One or more behavior factors are associated with abnormal cholesterol after adjustment  
 for demographic and body weight characteristics.
:::

# Initialize

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                        warning = TRUE,
                        message = TRUE,
                        fig.width = 8,
                        fig.height = 6
                        )
library(tidyr)
library(survey)
library(ggplot2)
library(srvyr)
library(dplyr)
library(haven)
library(janitor)
library(car)
```



# Define Data Directories
```{r define-data-directories}

project.dir <- "/Users/zhaozhan/nhanes-abnormal-cholesterol-young-adults"
data.dir <- file.path(project.dir, "01 Data")
output.dir <- file.path(project.dir, "02 Output")
processed.dir <- file.path(data.dir, "processed")


nhanes.young <- readRDS(
    file.path(processed.dir, "nhanes_young_cholesterol.rds")
)

# Questionaire files for RQ2 behavioral factors
alq.file <- file.path(data.dir, "ALQ_L.xpt")
paq.file <- file.path(data.dir, "PAQ_L.xpt")
slq.file <- file.path(data.dir, "SLQ_L.xpt")
smq.file <- file.path(data.dir, "SMQ_L.xpt")

alq <- read_xpt(alq.file) %>% clean_names()
paq <- read_xpt(paq.file) %>% clean_names()
slq <- read_xpt(slq.file) %>% clean_names()
smq <- read_xpt(smq.file) %>% clean_names()

# Check variable names
names(alq)
names(paq)
names(slq)
names(smq)

```



# Load and Clean Data 
```{r merge-data}
# Merge the behavioral data with the main dataset 
nhanes.q2 <- nhanes.young |>
    left_join(
        alq |> select(seqn, alq111, alq121, alq170), 
        by = "seqn") |>
    left_join(
        paq |> select(seqn, pad790q, pad790u, pad800, pad810q, pad810u, pad820, pad680), 
        by = "seqn") |>
    left_join(
        slq |> select(seqn, sld012, sld013),
        by = "seqn") |>
    left_join(
        smq |> select(seqn, smq020, smq040 ),
         by = "seqn")

```

# Create Analytic Variables and Survey Design

```{r create-analytic-variables}
nhanes.q2 <- nhanes.q2 |>

  mutate(

    # Outcome

    flag.any.num = as.numeric(flag.any),

    # Smoking status

    smoking_status = case_when(

      smq020 == 2 ~ "Never smoker",

      smq020 == 1 & smq040 == 3 ~ "Former smoker",

      smq020 == 1 & smq040 %in% c(1, 2) ~ "Current smoker",

      TRUE ~ NA_character_

    ),

    # Alcohol use

    alcohol_12mo = case_when(

      alq111 == 2 ~ "Never drinker",

      alq121 == 0 ~ "No alcohol in past 12 months",

      alq121 %in% 1:10 ~ "Alcohol use in past 12 months",

      TRUE ~ NA_character_

    ),

    binge_30d = case_when(

      alq170 == 0 ~ "No binge drinking in past 30 days",

      alq170 >= 1 & alq170 <= 30 ~ "Any binge drinking in past 30 days",

      TRUE ~ NA_character_

    ),

    # Clean physical activity special values

    pad790q_clean = if_else(pad790q %in% c(7777, 9999), NA_real_, as.numeric(pad790q)),

    pad800_clean  = if_else(pad800 %in% c(7777, 9999), NA_real_, as.numeric(pad800)),

    pad810q_clean = if_else(pad810q %in% c(7777, 9999), NA_real_, as.numeric(pad810q)),

    pad820_clean  = if_else(pad820 %in% c(7777, 9999), NA_real_, as.numeric(pad820)),

    pad680_clean  = if_else(pad680 %in% c(7777, 9999), NA_real_, as.numeric(pad680)),

    moderate_freq_week = case_when(

      pad790u == "D" ~ pad790q_clean * 7,

      pad790u == "W" ~ pad790q_clean,

      pad790u == "M" ~ pad790q_clean / 4.345,

      pad790u == "Y" ~ pad790q_clean / 52,

      TRUE ~ NA_real_

    ),

    vigorous_freq_week = case_when(

      pad810u == "D" ~ pad810q_clean * 7,

      pad810u == "W" ~ pad810q_clean,

      pad810u == "M" ~ pad810q_clean / 4.345,

      pad810u == "Y" ~ pad810q_clean / 52,

      TRUE ~ NA_real_

    ),

    moderate_min_week = moderate_freq_week * pad800_clean,

    vigorous_min_week = vigorous_freq_week * pad820_clean,

    total_pa_min_week = case_when(

      is.na(moderate_min_week) & is.na(vigorous_min_week) ~ NA_real_,

      TRUE ~ rowSums(cbind(moderate_min_week, vigorous_min_week), na.rm = TRUE)

    ),

    pa_category = case_when(

      total_pa_min_week >= 150 ~ "Meets 150 min/week",

      total_pa_min_week < 150 ~ "Below 150 min/week",

      TRUE ~ NA_character_

    ),

    sedentary_cat = case_when(

      pad680_clean >= 480 & pad680_clean < 1440 ~ "High sedentary time",

      pad680_clean < 480 ~ "Lower sedentary time",

      TRUE ~ NA_character_

    ),

    sleep_cat = case_when(
        
      sld012 < 7 ~ "Short sleep",

      sld012 >= 7 & sld012 <= 9 ~ "Recommended sleep",

      sld012 > 9 ~ "Long sleep",

      TRUE ~ NA_character_
    ),

    # Set reference categories
    smoking_status = factor(
      smoking_status,
      levels = c("Never smoker", "Former smoker", "Current smoker")
    ),
    binge_30d = factor(
      binge_30d,
      levels = c("No binge drinking in past 30 days", "Any binge drinking in past 30 days")
    ),
    pa_category = factor(
      pa_category,
      levels = c("Meets 150 min/week", "Below 150 min/week")
    ),
    sedentary_cat = factor(
      sedentary_cat,
      levels = c("Lower sedentary time", "High sedentary time")
    ),
    sex = factor(sex, levels = c("Female", "Male")),
    bmi.cat = factor(
      bmi.cat,
      levels = c("Underweight or normal weight", "Overweight", "Obesity")

    )

  )
options(survey.lonely.psu = "adjust")

nhanes.q2.design <- nhanes.q2 |>
  as_survey_design(
    ids = sdmvpsu,
    strata = sdmvstra,
    weights = wtmec2yr,
    nest = TRUE
  )

```

```{r descriptive-table}

# smoking status
q2.desc <- nhanes.q2.design |>
  group_by(smoking_status) |>
  summarise(
    prevalence = survey_mean(flag.any, na.rm = TRUE)
  ) |>
  mutate(
    prevalence_percent = sprintf("%.1f", prevalence * 100),
    se_percent = sprintf("%.1f", prevalence_se * 100)
  ) |>
  select(
    `Smoking status` = smoking_status,
    `Prevalence (%)` = prevalence_percent,
    `SE (%)` = se_percent
  )

q2.alcohol <- nhanes.q2.design |>
  group_by(binge_30d) |>
  summarise(
    prevalence = survey_mean(flag.any, na.rm = TRUE)
  ) |>
  mutate(
    prevalence_percent = sprintf("%.1f", prevalence * 100),
    se_percent = sprintf("%.1f", prevalence_se * 100)
  ) |>
  select(
    `Binge drinking` = binge_30d,
    `Prevalence (%)` = prevalence_percent,
    `SE (%)` = se_percent
  )

q2.pa <- nhanes.q2.design |>
  group_by(pa_category) |>
  summarise(
    prevalence = survey_mean(flag.any, na.rm = TRUE)
  ) |>
  mutate(
    prevalence_percent = sprintf("%.1f", prevalence * 100),
    se_percent = sprintf("%.1f", prevalence_se * 100)
  ) |>
  select(
    `Physical activity` = pa_category,
    `Prevalence (%)` = prevalence_percent,
    `SE (%)` = se_percent
  )

q2.sedentary <- nhanes.q2.design |>
  group_by(sedentary_cat) |>
  summarise(
    prevalence = survey_mean(flag.any, na.rm = TRUE)
  ) |>
  mutate(
    prevalence_percent = sprintf("%.1f", prevalence * 100),
    se_percent = sprintf("%.1f", prevalence_se * 100)
  ) |>
  select(
    `Sedentary behavior` = sedentary_cat,
    `Prevalence (%)` = prevalence_percent,
    `SE (%)` = se_percent
  )

knitr::kable(q2.desc, caption = "Table 1A. Weighted prevalence of abnormal cholesterol by smoking status.")

knitr::kable(q2.alcohol, caption = "Table 1B. Weighted prevalence of abnormal cholesterol by binge drinking status.")

knitr::kable(q2.pa, caption = "Table 1C. Weighted prevalence of abnormal cholesterol by physical activity status.")

knitr::kable(q2.sedentary, caption = "Table 1D. Weighted prevalence of abnormal cholesterol by sedentary behavior.")
```


Survey-weighted logistic regression models were used to examine associations between behavioral factors and abnormal cholesterol.  
The dependent variable was any abnormal cholesterol, defined as high total cholesterol, low HDL-C, or high non-HDL-C.  
Separate crude and adjusted models were fit for smoking status, binge drinking, physical activity, and sedentary behavior.  
Adjusted models included age, sex, race/ethnicity, and body weight status as covariates. Odds ratios and 95% confidence intervals were reported.

# Logistic Regression Models
```{r logistics-regression}
# Function to extract odds ratios and 95% confidence intervals

extract_or <- function(model) {
  est <- coef(model)
  ci <- confint(model)
  result <- data.frame(
    term = names(est),
    odds_ratio = exp(est),
    conf_low = exp(ci[, 1]),
    conf_high = exp(ci[, 2]),
    p_value = summary(model)$coefficients[, 4],
    row.names = NULL
  )
  result
}

# Crude model 1: smoking status
model_smoking_crude <- survey::svyglm(
  flag.any.num ~ smoking_status,
  design = nhanes.q2.design,
  family = quasibinomial()
)

or_smoking_crude <- extract_or(model_smoking_crude)
print(or_smoking_crude)


# Crude model 2: binge drinking
model_binge_crude <- survey::svyglm(
  flag.any.num ~ binge_30d,
  design = nhanes.q2.design,
  family = quasibinomial()
)

or_binge_crude <- extract_or(model_binge_crude)
print(or_binge_crude)


# Crude model 3: physical activity
model_pa_crude <- survey::svyglm(
  flag.any.num ~ pa_category,
  design = nhanes.q2.design,
  family = quasibinomial()
)

or_pa_crude <- extract_or(model_pa_crude)
print(or_pa_crude)


# Crude model 4: sedentary behavior
model_sedentary_crude <- survey::svyglm(
  flag.any.num ~ sedentary_cat,
  design = nhanes.q2.design,
  family = quasibinomial()
)

or_sedentary_crude <- extract_or(model_sedentary_crude)
print(or_sedentary_crude)

# Adjusted model 1: smoking status
model_smoking_adj <- survey::svyglm(
  flag.any.num ~ smoking_status + ridageyr + sex + factor(ridreth3) + bmi.cat,
  design = nhanes.q2.design,
  family = quasibinomial()
)

or_smoking_adj <- extract_or(model_smoking_adj)
print(or_smoking_adj)


# Adjusted model 2: binge drinking
model_binge_adj <- survey::svyglm(
  flag.any.num ~ binge_30d + ridageyr + sex + factor(ridreth3) + bmi.cat,
  design = nhanes.q2.design,
  family = quasibinomial()
)

or_binge_adj <- extract_or(model_binge_adj)
print(or_binge_adj)


# Adjusted model 3: physical activity
model_pa_adj <- survey::svyglm(
  flag.any.num ~ pa_category + ridageyr + sex + factor(ridreth3) + bmi.cat,
  design = nhanes.q2.design,
  family = quasibinomial()
)

or_pa_adj <- extract_or(model_pa_adj)
print(or_pa_adj)


# Adjusted model 4: sedentary behavior
model_sedentary_adj <- survey::svyglm(
  flag.any.num ~ sedentary_cat + ridageyr + sex + factor(ridreth3) + bmi.cat,
  design = nhanes.q2.design,
  family = quasibinomial()
)

or_sedentary_adj <- extract_or(model_sedentary_adj)
print(or_sedentary_adj)

rq2_adjusted_results <- bind_rows(
  or_smoking_adj |> mutate(model = "Smoking status"),
  or_binge_adj |> mutate(model = "Binge drinking"),
  or_pa_adj |> mutate(model = "Physical activity"),
  or_sedentary_adj |> mutate(model = "Sedentary behavior")
) |>
  select(model, term, odds_ratio, conf_low, conf_high, p_value) |>
  mutate(
    odds_ratio = round(odds_ratio, 2),
    conf_low = round(conf_low, 2),
    conf_high = round(conf_high, 2),
    p_value = round(p_value, 3)
  )

rq2_behavioral_results <- rq2_adjusted_results |>
  filter(
    term %in% c(
      "smoking_statusFormer smoker",
      "smoking_statusCurrent smoker",
      "binge_30dAny binge drinking in past 30 days",
      "pa_categoryBelow 150 min/week",
      "sedentary_catHigh sedentary time"
    )
  )

print(rq2_behavioral_results)

write.csv(
  rq2_behavioral_results,
  file = file.path(output.dir, "rq2_behavioral_main_results.csv"),
  row.names = FALSE
)
```

```{r rq2-kable-table, echo=FALSE, message=FALSE, warning=FALSE}

# Create a clean academic table for the main RQ2 behavioral results
rq2_behavioral_results <- rq2_adjusted_results |>
  filter(
    term %in% c(
      "smoking_statusFormer smoker",
      "smoking_statusCurrent smoker",
      "binge_30dAny binge drinking in past 30 days",
      "pa_categoryBelow 150 min/week",
      "sedentary_catHigh sedentary time"
    )
  )

rq2_behavioral_table <- rq2_behavioral_results |>
  mutate(
    behavioral_factor = case_when(
      term == "smoking_statusFormer smoker" ~ 
        "Former smoker vs. never smoker",
      term == "smoking_statusCurrent smoker" ~ 
        "Current smoker vs. never smoker",
      term == "binge_30dAny binge drinking in past 30 days" ~ 
        "Any binge drinking vs. none",
      term == "pa_categoryBelow 150 min/week" ~ 
        "Below 150 min/week physical activity vs. meets guideline",
      term == "sedentary_catHigh sedentary time" ~ 
        "High sedentary time vs. lower sedentary time",
      TRUE ~ term
    ),
    odds_ratio = sprintf("%.2f", odds_ratio),
    ci_95 = paste0(sprintf("%.2f", conf_low), "–", sprintf("%.2f", conf_high)),
    p_value = ifelse(p_value < 0.001, "<0.001", sprintf("%.3f", p_value))
  ) |>
  select(
    Model = model,
    `Behavioral Factor` = behavioral_factor,
    `Adjusted OR` = odds_ratio,
    `95% CI` = ci_95,
    `p-value` = p_value
  )

knitr::kable(
  rq2_behavioral_table,
  caption = "Table 2. Adjusted associations between behavioral factors and abnormal cholesterol among U.S. young adults aged 18–26 years, NHANES August 2021–August 2023.",
  align = c("l", "l", "c", "c", "c")
)


write.csv(
  rq2_behavioral_table,
  file = file.path(output.dir, "table2_rq2_behavioral_adjusted_or.csv"),
  row.names = FALSE
)

```

# logistic regression diagnostics
## Multicollinearity Check
```{r rq2-vif-check}
# Multicollinearity check using VIF
# Use a comparable unweighted glm for VIF screening

model_pa_adj_glm <- glm(
  flag.any.num ~ pa_category + ridageyr + sex + factor(ridreth3) + bmi.cat,
  data = nhanes.q2,
  family = binomial(),
  na.action = na.exclude
)

car::vif(model_pa_adj_glm)
```
Variance inflation factors were examined using an unweighted logistic regression model with the same covariates as a screening check for multicollinearity.  
The adjusted GVIF values were close to 1 for all predictors, suggesting no evidence of problematic multicollinearity among the model covariates.  

## Influence Diagnostics
```{r rq2-influence-check, fig.width=8, fig.height=5}
# Influence diagnostics using comparable unweighted logistic regression

model_pa_adj_glm <- glm(
  flag.any.num ~ pa_category + ridageyr + sex + factor(ridreth3) + bmi.cat,
  data = nhanes.q2,
  family = binomial(),
  na.action = na.exclude
)

diagnostic_data <- data.frame(
  obs_number = seq_along(dffits(model_pa_adj_glm)),
  dffits = dffits(model_pa_adj_glm)
)

cutoff <- 2 * sqrt(length(coef(model_pa_adj_glm)) / nobs(model_pa_adj_glm))

ggplot(diagnostic_data, aes(x = obs_number, y = dffits)) +
  geom_point() +
  geom_hline(yintercept = c(-cutoff, cutoff), linetype = "dashed") +
  labs(
    title = "DFFITS Diagnostic Plot",
    x = "Observation Number",
    y = "DFFITS"
  ) +
  theme_classic()
```
